# Missing values

My dataset is quite large, so for the purpose of visualization, I restrict the analysis to 2016, and assume that the pattern will be similar for years 2012 and 2008. Moreover, I deleted the variables which were derived from other variables, as their missing pattern would be exactly the same. I also deleted variables whose missingness is not meaningful (e.g. variable for additional notes, which is (a) optional and (b) text data).

```{r}
df16<-df.self %>%
  filter(year==2016) %>%
  select(-X,-case_id,-st,-dist,-pid3,-pid3_leaner,-turnout_selfreport,-state_po,
         -state_fips,-state_cen,-state_ic,-office,-Rpartyloss,-Rpartywins,-stable_resid_cons,
         -independent_strong,-independent_weak,-margin_003,-margin_002,-margin_001,
         -candidate,-party,-candidatevotes,-totalvotes,-notes,-share,-margin,-y_missing,
         -birthyr,-lostElection,-lossloss,-newsint_01,-newsint_012,-strong_partisan,-overallwin,-stable_resid, -church_reg, -church_ever,
         -overallwin,-year,-vep_highest_office,-version,-party7,-party_lean,-party3,-cd,-county_fips,-writein)

```

```{r}
#install.packages("mi")
#install.packages("betareg")
#install.packages("remotes")
library(dplyr)
library(mi)
library(betareg)
library(remotes)
remotes::install_github("cran/extracat")

# colSums(is.na(df16)) %>%
#   sort(decreasing= T) 

x<-missing_data.frame(df16)
image(x)

#x@patterns # commented out due to large data size.
#levels(x@patterns)
#summary(x@patterns)

extracat::visna(df16, sort = "c")

```

As one can see immediately from the charts, the only variable with a substantial number of missing values is the *turnout_self01*, i.e. self-reported turnout, which is the dependent variable of interest. The other variables with some missing data constitute such a small share of the sample that their aggregate measure barely appear on the plot. Given their negligible share, it would be easiest to drop these values from the analysis.

As for the missing turnout variable, given that this is the dependent variable of interest, I wouldn't feel comfortable imputing the value. As a result, I will drop the NAs by default in all illustrations unless otherwise indicated.