# Missing values

My dataset is quite large, so for the purpose of visualization I deleted the variables which were derived from other variables, as their missing pattern would be the same. I also deleted variables whose missingness is not meaningful (e.g. variable for additional notes, which is (a) optional and (b) text data). I also restrict the analysis to year 2016, which still leaves me with 5,783 observations.

```{r}
df16<-df.self %>%
  filter(year==2016) %>%
  select(-X,-case_id,-st,-dist,-pid3,-pid3_leaner,-turnout_selfreport,-state_po,
         -state_fips,-state_cen,-state_ic,-office,-Rpartyloss,-Rpartywins,-stable_resid_cons,
         -independent_strong,-independent_weak,-margin_003,-margin_002,-margin_001,
         -candidate,-party,-candidatevotes,-totalvotes,-notes,-share,-margin,-y_missing,
         -birthyr,-lostElection,-lossloss,-newsint_01,-newsint_012,-church1,-overallwin,-stable_resid,
         -overallwin,-year,-vep_highest_office,-version,-party7,-party_lean,-party3,-cd,-county_fips,-writein)

```

```{r,echo=FALSE}
#install.packages("mi")
#install.packages("betareg")
#install.packages("remotes")
library(mi)
library(betareg)
library(remotes)
library(dplyr)
remotes::install_github("cran/extracat")

colSums(is.na(df16)) %>%
  sort(decreasing= T) 

x<-missing_data.frame(df16)
image(x)

#x@patterns # commented out due to large data size.
levels(x@patterns)
summary(x@patterns)

extracat::visna(df16, sort = "c")

```

As one can see immediately from the charts, the only variable with a substantial number of missing values is the *turnout_self01*, i.e. self-reported turnout, which is the dependent variable of interest. Out of the 5,783 observations, 916 are missing this value. The variable that has the next highest number of values is *pid7* political affiliation using 7 scales, with only 9 missing values. Strong_partisanship is a variable that is perfectly aligned in terms of categories as pid7, so it also has 9 missing values. Again, there are 4 additional variables with missing values (newint, residence, faminc) but they are in the single digits and given that it's a neglibile share relative to the size of the full dataset, it would be easiest to drop these values from the analysis.

As for the missing turnout variable, given that this is the dependent variable of interest, I wouldn't feel comfortable imputing the value. As a result, I will drop the NAs.